---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---
## Loading necessary libraries
```{r global_options, include=TRUE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='figures/',
                      echo=TRUE, warning=FALSE, message=FALSE)
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(lattice))
```

## Loading and preprocessing the data
```{r, echo=TRUE}
unzip("activity.zip")
dat<-read.csv("activity.csv",header=TRUE)



totalsteps<- dat %>% group_by(date) %>% summarize(steps=sum(steps, na.rm=TRUE))

hist(totalsteps$steps, main="Histogram of Total Steps per Day", xlab="Total Steps")

```



## What is mean total number of steps taken per day?
```{r}
meansteps1<-mean(totalsteps$steps)
mediansteps1<-median(totalsteps$steps)
```

The mean total steps per day is `r meansteps1`, and the mediean total steps per day is `r mediansteps1`.

## What is the average daily activity pattern?
```{r}
interval<- dat %>% group_by(interval) %>% summarize(meansteps=mean(steps, na.rm=TRUE))
plot(interval$interval, interval$meansteps, type="l", main="Time Series Plot of Mean Steps by Interval", xlab="5 Minute Interval", ylab="Mean Steps")

maxsteps<-interval[which.max(interval$meansteps),]

```

The maximum average steps of `r maxsteps$meansteps` occur during interval `r maxsteps$interval`.


## Imputing missing values
```{r}
NAvalues<-sum(is.na(dat$steps))
```
The total number of observations missing steps is `r NAvalues`.

The mean steps for that interval was used to fill in the missing values. To do this, I joined the mean steps per interval to the main data frame, and substituting the mean steps per interval into the missing value using a subset function. 
```{r}
meansteps<-dat %>% group_by(interval) %>% summarize(meansteps=mean(steps, na.rm=TRUE))
datfill<-right_join(dat,meansteps)
datfill$steps[is.na(datfill$steps)]<-datfill$meansteps[is.na(datfill$steps)]
totalsteps<- datfill %>% group_by(date) %>% summarize(steps=sum(steps, na.rm=TRUE))
hist(totalsteps$steps, main="Histogram of Total Steps per Day", xlab="Total Steps")
meansteps2<-mean(totalsteps$steps)
mediansteps2<-median(totalsteps$steps)
meandiff<-meansteps2-meansteps1
mediandiff<-mediansteps2-mediansteps1
```
The mean total number of steps is `r meansteps2` and the median total number of steps is `r mediansteps2`. The difference between the mean steps is `r meandiff`. The difference between the median steps is `r mediandiff`.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
datfill<- datfill %>% transform(date=as.character(date))
datfill<- datfill %>% transform(date=as.character(date)) %>% transform(date=strptime(date,"%Y-%m-%d")) %>% transform(date=as.POSIXct(date))

datfill <- datfill %>% mutate(daytype=as.factor(ifelse(weekdays(datfill$date) %in% c("Saturday","Sunday"),"weekend","weekday")))

avgtypeint<-datfill %>% group_by(interval,daytype) %>% summarize(mean(steps))

names(avgtypeint)<-c("interval","daytype","avgsteps")

xyplot(avgsteps ~ interval|daytype, data=avgtypeint,layout=c(1,2),type="l",xlab="Five Minute Interval",ylab="Average Total Steps Taken")

```